The CMS Data Aggregation System

### 2010-05-04 doc/paper.tex
Fix spelling with ispell
git-svn-id: svn+ssh://svn.cern.ch/reps/CMSDMWM/DAS/trunk@665 4525493e-7705-40b1-a816-d608a930855b

%\cortext[cor1]{Corresponding autho}

@@

%% main text
\section{Introduction}
The European Organization for Nuclear Research, known as CERN, plays a leading
role in fundamental studies of physics. But it is also known as a place where
many innovations in the area of computer science were developed, e.g. World Wide Web.
Today, the Large Hadron Collider (LHC) at CERN is marking a new era of High Energy
Physics (HEP), promising to deliver a few PB of data each year. 
At this scale the ``information discovery'' within heterogeneous, distributed 
environment becomes a key ingredient of successful data analysis.
The data and associated meta-data are produced in variety of forms and digital formats.
They are stored and retrieved from relational and non-releational data-sources, such as 
RDMS systems, document oriented databases, blogs, twikies, file systems,
customized applications, etc. 
Working in such environment requires a lot of expertise and our users, physicists, 
are always looking for simple, intuitive and flexible
tool(s) to look-up their desired data. A well-known solutions, such as data-services
and search engines are tighted to a specific data sources and end-users are left 
with a manual task to bookkeep and relatate information from them.

@@

\section{Related Work\label{RelatedWork}}
Even though the idea of keyword queries over the relational database
%querying relation databases via keyword based search
%algorithms 
is not knew it is still under discussion in computer science domain.
A few alternative solutions has been proposed to address this issue
in last several years. In \cite{DBXplorer} the conjunctive keyword queries,
i.e. retrieval of documents that contain all query keywords, has
been discussed. It requires that all keywords are matched in a row from a
single table or joined tables. Authors of \cite{QueryAnswer} made a step forward
by considering information around matched values and presents customized answers to
end-users over the generated new database. These and other related 
works, see \cite{DBXplorer, QueryAnswer} for more references,
are based on some kind of the graph over the underlying database and
use a generated database out of pristine one to perform the work.
The proximity of the results in those techniques 
are mostly confined to a text-based values stored in a database, since 
usage of numerical values in the input search field is mostly useless without 
knowledge of the context of the value. For example, typing a value 100 
in the search input field leads to plenty of inappropriate matches, 
such as row ids, and in order to be
answered correctly requires additional keyword, clarifying its meaning, 
which should be matched with particular
entity (table.column) in the underlying database.
Therefore to answer presicely the logical questions, such as  
{\it find me total number of files whose size more then 20 but less then 
100} would be hard or almost impossible in \cite{DBXplorer, QueryAnswer}, 
while it can be easily accomplished using SQL syntax.

@@

\section{CMS data model\label{DataModel}}
The CMS, Compact Muon Solenoid, \cite{CMS} 
is one of the two large general-purpose particle physics detectors built on 
the proton-proton Large Hadron Collider (LHC) at CERN in Switzerland and France. 
It is designed to explore the frontier of the HEP and provide physicists
the ability to look at conditions presented at the early stage of our Universe.
More then 3000 physicists from 183 institutions reprepresenting almost 
38 countries are involved in the design, construction and maintenance of the experiment.

The CMS distributed computing and data model \cite{CMSDataModel} 
is designed to process
and efficiently manage a few PBs of data expected each year
in CMS operation at LHC. The computing resources provided by members of CMS
collaboration are geographically distributed, 
interconnected via high throughput networks and operated by means 
of Grid software. The model allows to cover broad variety of
hardware, mass and storage elements and configuration of the
clusters. To accomodate efficient data processing CMS uses
a multi-Tier distributed model, where specific tasks of data taking,
processing, archival and distribution are assigned to each tier based
on CMS data model. For example, the Tier-0 center at CERN is responsible
for archiving the data coming out from the detector, prompt first pass reconstruction
and data distribution to Tier-1 centers, located around the world.
%The 7 Tier-1 centers
%located in France, Germany, Ialy, Spain, Taiwan, United Kingdom and United States
%keep a portion (copy) of the data delivered by Tier-0 for further processing.
They provide storage and CPU power for high priority analysis jobs.
The Tier-2 centers
%, located at more then 50 sites around the world,
are dedicated for user analysis tasks and production of simulated data.

@@

\section{Data Aggregation System\label{DAS}}
The goal of the Data Aggregation System was to combine and
aggregate data from existing data-services
under one umrella, utilizing their access rules, APIs, security policies, etc., and
allowing users to perform queries against them. In contrast, the approaches
discussed in \cite{DBXplorer, QueryAnswer, FedDB} are
enforced a publication step of data into a dedicated database 
or central repositories \cite{iRODS}. Therefore, quite often it is required
that data should be converted into uniform data formats in order enable
aggregation, see \cite{OpenArchive}. In DAS architecture the data are provided
by existing data-services and infrastructure, whose implementation,
data representation and security policies are left intact. Consequently,
DAS does not require a data preservation and transaction capabilities. 
Instead DAS can be treated
as proxy or caching layer where information retrieval are triggered dynamically
by user queries. Such infrastructure allows data sharing across different queries and
data aggregation using appropriate mapping. Below we discuss in details the
DAS architecture.

\subsection{DAS architecture}
The design of the DAS system was based on previous studies of the CMS Data 
Bookkeeping System (DBS) \cite{DBS, DBS07}. The DBS Data Discovery service
\cite{DD} was based on DBS Query Language (DBS-QL) \cite{DBS-QL}. It provided
a syntax similar to SQL, without specific knowledge of underlying DB schema.
%We carefully analyzed user
%requests and queries placed into the DBS, their patterns, frequencies and latency. 
%As a result we created a data discovery service \cite{DD}. Originally it used MVC
%architecture and later, being re-factored as pure presentation layer on top of 
%the DBS Query Language (DBS-QL) \cite{DBS-QL} with support of dynamic queries against
%DBS back-ends. The DBS-QL \cite{DBS-QL} syntax was based on SQL query language, where 
%we took away relations and details on underlying schema from end-users. 
%This was achieved by using Dijkstras shortest path algorithm 
%to establish necessary table joins based on provided information in user queries and
%knowledge of foreign key relationships from auto-loaded database schema.
%The DBS-QL keys exposed to end-users were choosen based on a jargon used by
%physicists. They're internally mapped into database entitites and attributes, 
%e.g. the {\it run} DBS-QL key refers to run number attribute of run summary table.
A quick adoption and wide usage of DBS-QL in CMS gave us confidence in choosen approach 
and provide a basis for building DAS system.

As was pointed out in \cite{Arms} a mixed content and 
mixed meta-data and meta-data consistency should be considered as a whole in design 
of the system to successful information discovery. 
Starting from this ground we designed DAS system as an
additional layer on top of the existing data-services
within CMS computing infrastructure by imposing the following set of requirements:
\begin{itemize}
\item support a keyword based search queries with ability to use condition operators;
%\item pluggable interface to existing and up-coming data-service APIs;
\item use existing heterogeneous software environment and distributed nature of the data-services;
\item preserve access rules and security policies of individual data-services;
%\item be flexible to adjust to dynamic nature of data-service APIs, 
%due to continuously changing detector and software environments;
\item retrieve data on demand and aggregate them if necessary across
multiple data-services;
\item be transparent to data-services implementation and their data formats;
%\item support legacy data-services, applications and APIs.
\end{itemize}
\begin{figure}[htb]
\centering
\includegraphics[width=100mm]{DAS_Cache_and_Analytics.pdf}
\caption{
The DAS architecture diagram. 
Input query were placed into analytics DB and mapped into a set of
data-service APIs and their parameters. If superset of the query has been
found in DAS cache, the appropriate results were delivered back to the users.
Otherwise a data-service APIs were invoked, triggering DAS cache population. 
Once data become available and aggregated in DAS cache, users were able 
to see the results. This on demand system was supplemented by DAS robots 
(unix daemons) who were consulting DAS analytics database to enable data 
population into DAS cache for most common query requests.
}
\label{DAS_cache}
\end{figure}

\noindent
Based on this set we rule out the choice of RDMS as DAS back-end(s)
for several reasons. We agreed to support dynamic queries, data
aggregation on demand and pluggable data-services. Therefore a choice 
of concrete schema for data storage was problematic. Moreover we didn't impose
any transactions and data persistency in DAS. This allows us to look 
for alternative IT solutions.
Through the analysis of available options, such as a file-based and memory caches, 
key-value databases, documented-oriented databases we made our choice in favor 
of the last technology. Among them we evaluated CouchDB \cite{CouchDB} and 
MongoDB \cite{MongoDB}. Both systems provide schema free document
storage, replication and fail-over features, but our choice in favor of 
MongoDB was obvious, due to its support of dynamic queries, 
full indexies, including inner objects and embedded arrays,
and auto-sharding. Our prelimiary benchmarks shown that it can sustain
the desired load and size of our meta-data information. We used 
document-oriented database, MongoDB, as a back-end for the three 
main DAS components: Mapping, Analytics and DAS cache databases, 
which will be discussed next.

@@

\subsection{DAS core}
The DAS core communicated with different data-services and retrieved
data on demand. This was done by invoking appropriate data-service plugin(s)
upon the user request. Each plugin consists of DAS map, the
data parser(s) to convert data-service output into DAS records
and it was responsible for data-service API calls.
The data-service plugin contains a map about data-service APIs,
input and output parameters and their mapping to DAS keys. For example
\begin{verbatim}
{"api":{"params":{"name":""}, "name":"CMSNametoSE"}, "system":"sitedb",
 "api2das": [{"pattern": "re.compile('^T[0-3]_')", 
              "api_param": "name", "das_key": "site"}], 
 "daskeys": [{"pattern": "re.compile('^T[0-3]_')", 
              "key": "site", "map": "site.name"}]}
\end{verbatim}
This information was stored into DAS Mapping DB which we discuss next.

Although the majority of data-service plugins looks alike
and their logic was abstracted, including parsers, there were a few
cases where special care was required, e.g. access to data-services 
accessible via private network. Therefore, a way to propagate 
user credentials was required, leading to specific plugin implementation.

@@

\subsection{DAS Mapping DB}
The Mapping DB holded the mapping between data-service APIs and DAS keys exposed to
the users. It created a necessary glue between data-services and identifies
how their data can be related to each other. Basically Mapping DB was positioned 
as a {\it registration} service with ability to translate notations among data-services.
Upon identification of APIs which will participate in DAS activity 
we stored their names, input parameters, together with their type and accept 
values, as regular expression patterns, into the Mapping DB. 
Also we stored the mapping between data-service notations and DAS keys. 
For instance, the CMS DBS system \cite{DBS}
used a \verb+logical_file_name+ notation to identify the name of the
logical file, while the same entity was called \verb+lfn+ 
in CMS PhEDEx system \cite{PhEDEx}.
Since both of them were referred to the logical file name, 
we map both notations into DAS key \verb+file+.
%\footnote{We
%also deal with physical file names, but still used {\it file} to refer to them in
%DAS notations, since simple regular expressions were used to identify the
%meaning of the {\it file} keyword from the provided value.}
Such keys, e.g. \verb+file+, were exposed to the end-users as pre-defined keywords to
be used in search queries. For instance, end-users were able to type \verb+file+ in
order to get all file names known to DAS via data-services as well as
provide a concrete patterns associated with a file, e.g. \verb+file = abc*+.
Under the hood, the appropriate data-service APIs were looked up along with appropriate
parameters to answer those questions. In aforementioned example, 
usage of \verb+file+ in DAS query cause a call of DBS 
\verb+listFiles+ and PhEDEx \verb+fileReplicas+ APIs, respectively.
This approach gave ability to end-users to express the meaning of their queries, especially
when they deal with numbers, and achieve accuracy of results. 
For instance, users were able to type \verb+run > 10 and run < 100+
to specify a range of run numbers to be looked up in DAS cache. At the same time
a full keyword search is under development.
% within MongoDB.

@@

\subsection{DAS caching system}
The DAS caching system was used to dynamically fetch
and aggregate data upon user requests. We used the following 
algorithm:
\begin{enumerate}[1.]
\item Decompose query into set of conditions (\verb+conds+) 
and selection keys (\verb+skeys+)
\item Check cache for query hash, if found, look-up results from the cache
\item Identify list of APIs which can serve requested data
\begin{verbatim}
apilist = [r(url, api, args) for r in mapping_db(skeys, conds)]
\end{verbatim}
Here \verb+mappind_db+ defines translation between input query and data-service
APIs which can delivery such data.
\item Iterate over APIs
\begin{verbatim}
for url, api, args in apilist:
    if  superset_in_analytics(api, args):
        continue
    for row in parser(get_data(url, api, args)):
        yield row
\end{verbatim}
Here \verb+superset_in_analytics+ verifies if provided \verb+api, args+
are subset of already registered in Analytics DB api call(s). The 
\verb+get_data+ and \verb+parser+ are define methods to fetch and parse
data, respectively.
\item Perform aggregation by iterating over records from the previos step
\begin{verbatim}
if cache_has(row.primary_key):
   existing_record = get_from_cache(row.primary_key)
   merge(row, existing_record)
   insert_into_cache(row)
   delete_from_cache(existing_record)
else:
   insert_into_cache(row)
\end{verbatim}
The \verb+primary_key+
is a key which identify the record defined in Mapping DB, e.g. \verb+file.name+
for \verb+file+ record.
We did use a bulk insert operations at this step to increase throughput. 
\item Get results from the cache
\end{enumerate}
%The data was stored in a form of DAS records into the cache, while
%queries were recorded into Analytics DB for further analysis.
We also had an option to run independent set of daemons, DAS robots, who
monitor Analytics DB for most popular requests and pre-fetch the most
common data into DAS cache. 
%The interesting part was being able
%to identify the superset of api/args out of found one in an Analytics DB.

@@

\subsection{DAS queries}
Apart from the previous DBS-QL syntax we decided to drop its syntax and allow users
express their queries in a free text-based form. Although we must clarify its
meaning. We used a pre-defined set of keys registered in DAS to describe common
entities of our data. For example, \verb+dataset+, 
\verb+file+, \verb+block+, \verb+run+, \verb+event+. Those
names were used by our users in their daily work operations. So, by typing a
\verb+file+ we didn't perform a string matching to the 
``file'' string, but rather look-up all file objects in a DAS. 
Moreover each DAS key has a default attribute.
For intance, the default attribute for
\verb+dataset+, \verb+block+, \verb+file+ 
was a \verb+name+, while the default attribute for
\verb+run+ and \verb+event+ was a \verb+number+. 
This allows to specify key-value pair as simple as possible, e.g.
\verb+run=123456+, which means run number 123456. At the same time a query 
key, e.g. \verb+site+ meant to look-up all site objects. Such ability
to use key-value pairs resolve ambiguity in object values. For instance,
specifying a run number as \verb+run=123456+ instructs system to resolve
data object with run equal to 123456 rather then other objects which may have
123456 values, e.g. event number or file size. At the same time, to complement
the stregth of the DAS we do plan to index data objects and add a keyword-based 
search later.

@@

\subsection{Data-services and aggregation}
As we outlined above the DAS Mapping DB becomes an authorative
source of information about all data-services participating in DAS.
Based on information stored in Mapping DB we were able to 
retrieve appropriate data from different data-services, re-map them into 
DAS notations and aggregate them on demand. For instance, the Data 
Bookkeeping System (DBS) and data location system (PhEDEx)
provided information about CMS data block objects. In former case, DBS 
stored information about blocks and their relationships to other 
physics objects, such as dataset adn files, while PhEDEx kept
information about its location and replicas. Upon a user request to 
find a block, we were able to identify appropriate set of DBS and PhEDEx APIs, 
retrieve data from them, re-format their output to DAS notations 
and store results into DAS cache performing aggregation of matched 
objects using \verb+block.name+ key. The merged results were stored 
as a set of data records. Here is an example of such aggregated record:
\begin{verbatim}
{"das_id": ["4b195404e2194e21c4000030", "4b1953fee2194e21c4000002"], 
 "block": [{"name": "/a/b/c#123", "replica": [{"complete": "y", 
            "subscribed": "n", "custodial": "n", "site": "T1_US_Buffer", 
            "nfiles": "18", "se": "srm.site.com", 
            "size": "112694171838"}, {....}]},
           {"name": "a/b/c#123", "nevents": "570200", 
            "created_by": "USER_DN", "path": "/a/b/c", ...}],
 "das": [{"expire": 1259952908}, {"expire": 1259954702}]}
\end{verbatim}
Here \verb+das_id+ refers to api call records stored in Analytics DB,
\verb+block+ represents an aggregation block object information from DBS
and PhEDEx systems and \verb+das+ contains their expiration time stamps.
Usage of this information was valuable contribution in debugging process of
data-services themselves, e.g. identification of not-matched records, 
latency studies, etc.
% Here is an example of DAS record representing
%aggregation information between two SiteDB APIs, {\it CMStoSAMName} 
%and {\it CMStoSiteName}, with API results merged under the {\it site} key.
%\begin{verbatim}
%{"site":[{"samname": "CERN-PROD", "name": "T1_CH_CERN"}, 
%         {"sitename": "CERN", "name": "T1_CH_CERN"}], 
% "das": [{"ctime": 0.9, "expire": 1256362175, "api": ["CMStoSAMName"],
%          "url": "https://a.b.com/sitedb/json/index/CMStoSAMName", 
%          "timestamp": 1256318975, "system": ["sitedb"]},
%         {"ctime": 1.2, "api": ["CMStoSiteName"], "expire": 1256362174},
%          "url": "https://a.b.com/sitedb/json/index/CMStoSiteName", 
%          "timestamp": 1256318974, "system": ["sitedb"]]}
%\end{verbatim}

%We used RESTful model \cite{REST} for all DAS cache server APIs.
%It allows us to share core library and other components during development of 
%various DAS clients, e.g. CLI, web interface, robots, etc. and hides
%complexity of access to individual data-services and their security policies
%as well as proper organize caching solution.
%For instance, to access on-line Run Summary information we were forced to use
%Grid certificates. Instead of propagating user certificate to data-service
%the DAS cache server works as a proxy to this server and used host certficiate 
%to serve such requests on behalf of the users.

>>>>

%\cortext[cor1]{Corresponding author}

@@

%% main text
\section{Introduction}
The European Organization for Nuclear Research, known as CERN, plays a leading
role in fundamental studies of physics. But it is also known as a place where
many innovations in the area of computer science were developed, e.g. World Wide Web.
Today, the Large Hadron Collider (LHC) at CERN is marking a new era of High Energy
Physics (HEP), promising to deliver a few PB of data each year. 
At this scale the ``information discovery'' within heterogeneous, distributed 
environment becomes a key ingredient of successful data analysis.
The data and associated meta-data are produced in variety of forms and digital formats.
They are stored and retrieved from relational and non-relational data-sources, such as 
RDMS systems, document oriented databases, blogs, twikies, file systems,
customized applications, etc. 
Working in such environment requires a lot of expertise and our users, physicists, 
are always looking for simple, intuitive and flexible
tool(s) to look-up their desired data. A well-known solutions, such as data-services
and search engines are tighten to a specific data sources and end-users are left 
with a manual task to bookkeep and relate information from them.

@@

\section{Related Work\label{RelatedWork}}
Even though the idea of keyword queries over the relational database
%querying relation databases via keyword based search
%algorithms 
is not knew it is still under discussion in computer science domain.
A few alternative solutions has been proposed to address this issue
in last several years. In \cite{DBXplorer} the conjunctive keyword queries,
i.e. retrieval of documents that contain all query keywords, has
been discussed. It requires that all keywords are matched in a row from a
single table or joined tables. Authors of \cite{QueryAnswer} made a step forward
by considering information around matched values and presents customized answers to
end-users over the generated new database. These and other related 
works, see \cite{DBXplorer, QueryAnswer} for more references,
are based on some kind of the graph over the underlying database and
use a generated database out of pristine one to perform the work.
The proximity of the results in those techniques 
are mostly confined to a text-based values stored in a database, since 
usage of numerical values in the input search field is mostly useless without 
knowledge of the context of the value. For example, typing a value 100 
in the search input field leads to plenty of inappropriate matches, 
such as row ids, and in order to be
answered correctly requires additional keyword, clarifying its meaning, 
which should be matched with particular
entity (table.column) in the underlying database.
Therefore to answer precisely the logical questions, such as  
{\it find me total number of files whose size more then 20 but less then 
100} would be hard or almost impossible in \cite{DBXplorer, QueryAnswer}, 
while it can be easily accomplished using SQL syntax.

@@

\section{CMS data model\label{DataModel}}
The CMS, Compact Muon Solenoid, \cite{CMS} 
is one of the two large general-purpose particle physics detectors built on 
the proton-proton Large Hadron Collider (LHC) at CERN in Switzerland and France. 
It is designed to explore the frontier of the HEP and provide physicists
the ability to look at conditions presented at the early stage of our Universe.
More then 3000 physicists from 183 institutions representing almost 
38 countries are involved in the design, construction and maintenance of the experiment.

The CMS distributed computing and data model \cite{CMSDataModel} 
is designed to process
and efficiently manage a few PBs of data expected each year
in CMS operation at LHC. The computing resources provided by members of CMS
collaboration are geographically distributed, 
interconnected via high throughput networks and operated by means 
of Grid software. The model allows to cover broad variety of
hardware, mass and storage elements and configuration of the
clusters. To accommodate efficient data processing CMS uses
a multi-Tier distributed model, where specific tasks of data taking,
processing, archival and distribution are assigned to each tier based
on CMS data model. For example, the Tier-0 center at CERN is responsible
for archiving the data coming out from the detector, prompt first pass reconstruction
and data distribution to Tier-1 centers, located around the world.
%The 7 Tier-1 centers
%located in France, Germany, Italy, Spain, Taiwan, United Kingdom and United States
%keep a portion (copy) of the data delivered by Tier-0 for further processing.
They provide storage and CPU power for high priority analysis jobs.
The Tier-2 centers
%, located at more then 50 sites around the world,
are dedicated for user analysis tasks and production of simulated data.

@@

\section{Data Aggregation System\label{DAS}}
The goal of the Data Aggregation System was to combine and
aggregate meta-data from existing data-services
under one umbrella, utilizing their access rules, APIs, security policies, etc., and
allowing users to perform queries against them. In contrast, the approaches
discussed in \cite{DBXplorer, QueryAnswer, FedDB} are
enforced a publication step of data into a dedicated database 
or central repositories \cite{iRODS}. Therefore, quite often it is required
that data should be converted into uniform data formats in order enable
aggregation, see \cite{OpenArchive}. In DAS architecture the data are provided
by existing data-services and infrastructure, whose implementation,
data representation and security policies are left intact. Consequently,
DAS does not require a data preservation and transaction capabilities. 
Instead DAS can be treated
as proxy or caching layer where information retrieval are triggered dynamically
by user queries. Such infrastructure allows data sharing across different queries and
data aggregation using appropriate mapping. Below we discuss in details the
DAS architecture.

\subsection{DAS architecture}
The design of the DAS system was based on previous studies of the CMS Data 
Bookkeeping System (DBS) \cite{DBS, DBS07}. The DBS Data Discovery service
\cite{DD} was based on DBS Query Language (DBS-QL) \cite{DBS-QL}. It provided
a syntax similar to SQL, without specific knowledge of underlying DB schema.
%We carefully analyzed user
%requests and queries placed into the DBS, their patterns, frequencies and latency. 
%As a result we created a data discovery service \cite{DD}. Originally it used MVC
%architecture and later, being re-factored as pure presentation layer on top of 
%the DBS Query Language (DBS-QL) \cite{DBS-QL} with support of dynamic queries against
%DBS back-ends. The DBS-QL \cite{DBS-QL} syntax was based on SQL query language, where 
%we took away relations and details on underlying schema from end-users. 
%This was achieved by using Dijkstra's shortest path algorithm 
%to establish necessary table joins based on provided information in user queries and
%knowledge of foreign key relationships from auto-loaded database schema.
%The DBS-QL keys exposed to end-users were chosen based on a jargon used by
%physicists. They're internally mapped into database entities and attributes, 
%e.g. the {\it run} DBS-QL key refers to run number attribute of run summary table.
A quick adoption and wide usage of DBS-QL in CMS gave us confidence in chosen approach 
and provide a basis for building DAS system.

As was pointed out in \cite{Arms} a mixed content and 
mixed meta-data and meta-data consistency should be considered as a whole in design 
of the system to successful information discovery. 
Starting from this ground we designed DAS system as an
additional layer on top of the existing data-services
within CMS computing infrastructure by imposing the following set of requirements:
\begin{itemize}
\item support a keyword based search queries with ability to use condition operators;
%\item pluggable interface to existing and up-coming data-service APIs;
\item use existing heterogeneous software environment and distributed nature of the data-services;
\item preserve access rules and security policies of individual data-services;
%\item be flexible to adjust to dynamic nature of data-service APIs, 
%due to continuously changing detector and software environments;
\item retrieve data on demand and aggregate them if necessary across
multiple data-services;
\item be transparent to data-services implementation and their data formats;
%\item support legacy data-services, applications and APIs.
\end{itemize}
\begin{figure}[htb]
\centering
\includegraphics[width=100mm]{DAS_Cache_and_Analytics.pdf}
\caption{
The DAS architecture diagram. 
Input query were placed into analytics DB and mapped into a set of
data-service APIs and their parameters. If superset of the query has been
found in DAS cache, the appropriate results were delivered back to the users.
Otherwise a data-service APIs were invoked, triggering DAS cache population. 
Once data become available and aggregated in DAS cache, users were able 
to see the results. This on demand system was supplemented by DAS robots 
(Unix daemons) who were consulting DAS analytics database to enable data 
population into DAS cache for most common query requests.
}
\label{DAS_cache}
\end{figure}

\noindent
Based on this set we rule out the choice of RDMS as DAS back-end(s)
for several reasons. We agreed to support dynamic queries, data
aggregation on demand and pluggable data-services. Therefore a choice 
of concrete schema for data storage was problematic. Moreover we didn't impose
any transactions and data persistence in DAS. This allows us to look 
for alternative IT solutions.
Through the analysis of available options, such as a file-based and memory caches, 
key-value databases, documented-oriented databases we made our choice in favor 
of the last technology. Among them we evaluated CouchDB \cite{CouchDB} and 
MongoDB \cite{MongoDB}. Both systems provide schema free document
storage, replication and fail-over features, but our choice in favor of 
MongoDB was obvious, due to its support of dynamic queries, 
full indexes, including inner objects and embedded arrays,
and auto-sharding. Our preliminary benchmarks shown that it can sustain
the desired load and size of our meta-data information. We used 
document-oriented database, MongoDB, as a back-end for the three 
main DAS components: Mapping, Analytics and DAS cache databases, 
which will be discussed next.

@@

\subsection{DAS core}
The DAS core communicated with different data-services and retrieved
data on demand. This was done by invoking appropriate data-service plug-in(s)
upon the user request. Each plug-in consists of DAS map, the
data parser(s) to convert data-service output into DAS records
and it was responsible for data-service API calls.
The data-service plug-in contains a map about data-service APIs,
input and output parameters and their mapping to DAS keys. For example
\begin{verbatim}
{"api":{"params":{"name":""}, "name":"CMSNametoSE"}, "system":"sitedb",
 "api2das": [{"pattern": "re.compile('^T[0-3]_')", 
              "api_param": "name", "das_key": "site"}], 
 "daskeys": [{"pattern": "re.compile('^T[0-3]_')", 
              "key": "site", "map": "site.name"}]}
\end{verbatim}
This information was stored into DAS Mapping DB which we discuss next.

Although the majority of data-service plug-ins looks alike
and their logic was abstracted, including parsers, there were a few
cases where special care was required, e.g. access to data-services 
accessible via private network. Therefore, a way to propagate 
user credentials was required, leading to specific plug-in implementation.

@@

\subsection{DAS Mapping DB}
The Mapping DB hold the mapping between data-service APIs and DAS keys exposed to
the users. It created a necessary glue between data-services and identifies
how their data can be related to each other. Basically Mapping DB was positioned 
as a {\it registration} service with ability to translate notations among data-services.
Upon identification of APIs which will participate in DAS activity 
we stored their names, input parameters, together with their type and accept 
values, as regular expression patterns, into the Mapping DB. 
Also we stored the mapping between data-service notations and DAS keys. 
For instance, the CMS DBS system \cite{DBS}
used a \verb+logical_file_name+ notation to identify the name of the
logical file, while the same entity was called \verb+lfn+ 
in CMS PhEDEx system \cite{PhEDEx}.
Since both of them were referred to the logical file name, 
we map both notations into DAS key \verb+file+.
%\footnote{We
%also deal with physical file names, but still used {\it file} to refer to them in
%DAS notations, since simple regular expressions were used to identify the
%meaning of the {\it file} keyword from the provided value.}
Such keys, e.g. \verb+file+, were exposed to the end-users as pre-defined keywords to
be used in search queries. For instance, end-users were able to type \verb+file+ in
order to get all file names known to DAS via data-services as well as
provide a concrete patterns associated with a file, e.g. \verb+file = abc*+.
Under the hood, the appropriate data-service APIs were looked up along with appropriate
parameters to answer those questions. In aforementioned example, 
usage of \verb+file+ in DAS query cause a call of DBS 
\verb+listFiles+ and PhEDEx \verb+fileReplicas+ APIs, respectively.
This approach gave ability to end-users to express the meaning of their queries, especially
when they deal with numbers, and achieve accuracy of results. 
For instance, users were able to type \verb+run > 10 and run < 100+
to specify a range of run numbers to be looked up in DAS cache. At the same time
a full keyword search is under development.
% within MongoDB.

@@

\subsection{DAS caching system}
The DAS caching system was used to dynamically fetch
and aggregate data upon user requests. We used the following 
algorithm:
\begin{enumerate}[1.]
\item Decompose query into set of conditions (\verb+conds+) 
and selection keys (\verb+skeys+)
\item Check cache for query hash, if found, look-up results from the cache
\item Identify list of APIs which can serve requested data
\begin{verbatim}
apilist = [r(url, api, args) for r in mapping_db(skeys, conds)]
\end{verbatim}
Here \verb+mappind_db+ defines translation between input query and data-service
APIs which can delivery such data.
\item Iterate over APIs
\begin{verbatim}
for url, api, args in apilist:
    if  superset_in_analytics(api, args):
        continue
    for row in parser(get_data(url, api, args)):
        yield row
\end{verbatim}
Here \verb+superset_in_analytics+ verifies if provided \verb+api, args+
are subset of already registered in Analytics DB api call(s). The 
\verb+get_data+ and \verb+parser+ are define methods to fetch and parse
data, respectively.
\item Perform aggregation by iterating over records from the previous step
\begin{verbatim}
if cache_has(row.primary_key):
   existing_record = get_from_cache(row.primary_key)
   merge(row, existing_record)
   insert_into_cache(row)
   delete_from_cache(existing_record)
else:
   insert_into_cache(row)
\end{verbatim}
The \verb+primary_key+
is a key which identify the record defined in Mapping DB, e.g. \verb+file.name+
for \verb+file+ record.
We did use a bulk insert operations at this step to increase throughput. 
\item Get results from the cache
\end{enumerate}
%The data was stored in a form of DAS records into the cache, while
%queries were recorded into Analytics DB for further analysis.
We also had an option to run independent set of daemons, DAS robots, who
monitor Analytics DB for most popular requests and pre-fetch the most
common data into DAS cache. 
%The interesting part was being able
%to identify the superset of api/args out of found one in an Analytics DB.

@@

\subsection{DAS queries}
Apart from the previous DBS-QL syntax we decided to drop its syntax and allow users
express their queries in a free text-based form. Although we must clarify its
meaning. We used a pre-defined set of keys registered in DAS to describe common
entities of our data. For example, \verb+dataset+, 
\verb+file+, \verb+block+, \verb+run+, \verb+event+, etc. Their choice
was based on analysis of daily work operations of our users, see \cite{DBS07}.
Each key had a set of attributes with primary one assigned to the key itself.
For instance, the \verb+block+ key had the following attributes
\verb+name+, \verb+size+, \verb+replica+ etc., 
with primary attribute being \verb+name+.
The choice of attributes were based on a set of meta-data information
associated with a given DAS key. Therefore end-users where able to specify
a context in their queries by using appropriate attribute(s). For example,
for query \verb+block+ we didn't perform a string matching to the
``block'' string, but rather look-up a block objects in a DAS, while query
\verb+block.name=abc*+ instructs system to look-up appropriate blocks
with matched names. At the same time, to complement the strength of the 
DAS, we do plan to index dynamically retrieved meta-data and add a keyword-based 
search later.

@@

\subsection{Data-services and aggregation}
As we outlined above the DAS Mapping DB becomes an authorative
source of information about all data-services participating in DAS.
Based on information stored in Mapping DB we were able to 
retrieve appropriate data from different data-services, re-map them into 
DAS notations and aggregate them on demand. For instance, the Data 
Bookkeeping System (DBS) and data location system (PhEDEx)
provided information about CMS block meta-data. In former case, DBS 
stored information about blocks and their relationships to other 
physics objects, such as dataset and files, while PhEDEx kept
information about its location and replicas. Upon a user request to 
find a block meta-data, we were able to identify appropriate set of DBS and PhEDEx APIs, 
retrieve data from them, re-format their output to DAS notations 
and store results into DAS cache performing aggregation of matched 
objects using \verb+block.name+ key. The merged results were stored 
as a set of data records. Here is an example of such aggregated record:
\begin{verbatim}
{"das_id": ["4b195404e2194e21c4000030", "4b1953fee2194e21c4000002"], 
 "block": [{"name": "/a/b/c#123", "replica": [{"complete": "y", 
            "subscribed": "n", "custodial": "n", "site": "T1_US_Buffer", 
            "nfiles": "18", "se": "srm.site.com", 
            "size": "112694171838"}, {....}]},
           {"name": "a/b/c#123", "nevents": "570200", 
            "created_by": "USER_DN", "path": "/a/b/c", ...}],
 "das": [{"expire": 1259952908}, {"expire": 1259954702}]}
\end{verbatim}
Here \verb+das_id+ refers to api call records stored in Analytics DB,
\verb+block+ represents an aggregation block meta-data information from DBS
and PhEDEx systems and \verb+das+ contains their expiration time stamps.
Usage of this information was valuable contribution in debugging process of
data-services themselves, e.g. identification of not-matched records, 
latency studies, etc.
% Here is an example of DAS record representing
%aggregation information between two SiteDB APIs, {\it CMStoSAMName} 
%and {\it CMStoSiteName}, with API results merged under the {\it site} key.
%\begin{verbatim}
%{"site":[{"samname": "CERN-PROD", "name": "T1_CH_CERN"}, 
%         {"sitename": "CERN", "name": "T1_CH_CERN"}], 
% "das": [{"ctime": 0.9, "expire": 1256362175, "api": ["CMStoSAMName"],
%          "url": "https://a.b.com/sitedb/json/index/CMStoSAMName", 
%          "timestamp": 1256318975, "system": ["sitedb"]},
%         {"ctime": 1.2, "api": ["CMStoSiteName"], "expire": 1256362174},
%          "url": "https://a.b.com/sitedb/json/index/CMStoSiteName", 
%          "timestamp": 1256318974, "system": ["sitedb"]]}
%\end{verbatim}

%We used RESTful model \cite{REST} for all DAS cache server APIs.
%It allows us to share core library and other components during development of 
%various DAS clients, e.g. CLI, web interface, robots, etc. and hides
%complexity of access to individual data-services and their security policies
%as well as proper organize caching solution.
%For instance, to access on-line Run Summary information we were forced to use
%Grid certificates. Instead of propagating user certificate to data-service
%the DAS cache server works as a proxy to this server and used host certificate 
%to serve such requests on behalf of the users.
